### 2019-08-19 ThesisJo/2_background.tex
2019-08-18

\newif\ifTODO\TODOfalse                        % Use todo notes?

%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
% DOCUMENTCLASS %                                See full option description in "mytemplate.cls"
%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%

\documentclass[
a4paper,10pt
]{common/ica2013_2}

% \usepackage[utf8x]{inputenc}
\usepackage[table,dvipsnames,svgnames]{xcolor}
\usepackage{titlesec}

\newcommand\Fig[1]{Fig. \ref{#1}}

\newcommand\Grey[1]{{\color{Grey}#1}}
\newcommand\Red[1]{{\color{Red}#1}}
\newcommand\Blue[1]{{\color{Blue}#1}}
\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
\newcommand\Brown[1]{{\color{Brown}#1}}
\newcommand\Green[1]{{\color{Green}#1}}
\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
\newcommand\Yellow[1]{{\color{yellow}#1}}
\newcommand\Orange[1]{{\color{orange}#1}}

\ifTODO
   \newcounter{todoidx}
   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
   \setlength\marginparsep{1pt}
   \setlength\marginparwidth{50pt}
   \newlength\marginparwidthsmall
   \setlength\marginparwidthsmall{\marginparwidth}
   \addtolength\marginparwidthsmall{-7pt}
   \newcommand\todo[1]{%
      \addtocounter{todoidx}{1}%
      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
      \marginpar{%
         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}

   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
\else
%    \usepackage[disable]{./todonotes} 
   \newcommand\todo[1]{}
\fi

\newcommand\nn{\nonumber\\}

\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
\newcommand\case[1]{\begin{cases}#1\end{cases}}
\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}

\newcommand\rand{\text{rand}}
\newcommand\randn{\text{randn}}
\newcommand\rect{\text{rect}}
\newcommand\sinc{\text{sinc}}
\newcommand\tr{\text{tr}}
\newcommand\adj{\text{adj}}

% \newcommand\max{\text{max}}
\newcommand\argmin[1]{\text{arg}\,\underset{#1}{\text{min}}}

\newcommand\qqquad{\quad\qquad}
\newcommand\qqqquad{\qquad\qquad}

\renewcommand\l[1]{\left#1}
\renewcommand\r[1]{\right#1}

% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}

%Keyword colouring:
\newcommand\kw[1]{#1}
\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}

\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
\newcommand\var[3]{#1_\text{#2}\of{#3}}

\newcommand\diag{\text{diag}}

% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
\newcommand\mathnorm{\textstyle}
\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}

\newcommand\sumu[1]{\sum\limits^{#1}\,}
\newcommand\suml[1]{\sum\limits_{#1}\,}
\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\,}

\newcommand\produ[1]{\prod\limits^{#1}\,}
\newcommand\prodl[1]{\prod\limits_{#1}\,}
\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\,}

\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}

%Math macros:
\newcommand\diff[2]{\frac{\kw{d}\,\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
\newcommand\ddiff[2]{\frac{\kw{d^2}\,\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}

\renewcommand\d[1]{\scriptstyle\kw{\,d\parm{#1}}\displaystyle}

% These commands are mutually exclusive. Remember to "renew" in v2.
\newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
\newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
\newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}

\newcommand\T{^{\scriptscriptstyle T}}
\renewcommand\H{^{\scriptscriptstyle H}}

\renewcommand\vec[1]{\boldsymbol{#1}}
\newcommand\mat[1]{\boldsymbol{#1}}

\newcommand\I{\mat I}
\renewcommand*\a{\vec a}
\renewcommand*\i{\vec i}
\renewcommand*\k{\vec k}
\newcommand*\n{\vec n}
\newcommand*\p{\vec p}
\newcommand*\s{\vec s}
\newcommand*\w{\vec w}
\newcommand*\x{\vec x}
\newcommand*\y{\vec y}

\newcommand*\A{\mat A}
\newcommand*\B{\mat B}
\newcommand*\C{\mat C}
\newcommand*\E{\mat E}
% \renewcommand*\H{\mat H}
\renewcommand*\P{\mat P}
\newcommand*\eP{\mat{\hat P}}
\newcommand*\R{\mat R}
\newcommand*\Ri{\R^{-1}}
\newcommand*\eR{\mat{\hat R}}
\newcommand*\eRi{\hat{\mat R}\,\!^{-1}}
\newcommand*\Navg{N_\text{avg}}
\newcommand*\W{\mat W}
\newcommand*\X{\mat X}
\newcommand*\Xd{\X_{\!\Delta}}
\newcommand*\Y{\mat Y}
\renewcommand*\B{\mat B}

\renewcommand*\L{\mat \Lambda}
\newcommand*\U{\mat U}
% \renewcommand*\t{\mathtiny{^T}}
% \newcommand*\h{\mathtiny{^H}}
\renewcommand*\t{^T}
\newcommand*\h{^H}

\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
\newcommand\Dx{\vec\nabla\times}
\newcommand\Dd{\vec\nabla\cdot}

\usepackage{tikz}
\usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}
% \usepackage{glossaries}

\newcommand\graphicsAI[2][]{%
  \immediate\write18{./bin/laFigure #2 #1}%
  \input{result}}%

\definecolor{tabBlue}{HTML}{AACCFF}

\newenvironment{outline}
{\begin{itemize}}
{\end{itemize}}

\begin{document}
% \setlength{\headrulewidth}{0.0pt}

%~~ TitlePage ~~%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
\thispagestyle{empty}
\pagenumbering{arabic} %Normal numbers

{\Large\bf Adapting the MVDR beamformer to a GPU for differently sized active sonar imaging systems.}

\begin{abstract}
The MVDR beamformer has been shown to improve active sonar image quality compared to conventional methods. Unfortunately, it is also significantly more computationally expensive because a spatial covariance matrix must be estimated and inverted for each image pixel. We target this challenge by altering and mapping the MVDR beamformer to a GPU, and suggest three different solutions depending on the system size.

For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield image creation rates of more than 1\,Mpx/s. For larger systems we show that frequency domain processing is preferable, as this promotes high processing rates at a negligible reduction in image quality. These GPU implementations consistently reduced the runtime by 2-3 orders of magnitude compared to our reference C and Matlab implementations.

For even larger systems we suggest employing the LCA beamformer. It does not calculate a weightset, but merely computes the beamformer output for each of a predefined set of weights, and selects the one that best fulfils the MVDR criterion. The LCA creates images with a quality comparable to MVDR, and it is perfectly suited for a GPU.
\end{abstract}
\vspace{2cm}
(Forsiden genereres av ASA.)
% \ \\
% Total: 195 words (200 max)
% 
% \ \\
% Noen tanker jeg har selv naa:
% \begin{itemize}
% \item Det mangler noe om bruksomraader.
% \item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haaplos ut(?)
% \end{itemize}
% 
% 
% \ \\
% Questions they wanted answered:
% \begin{itemize}
% \item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
% \item If algorithm is slow - What technology will help me?
% \item What skills are needed?
% \item What tools are available and what's the cost?
% \item What acceleration factor are we looking at?
% \item How can I get started?
% \item What we did. Which tools? Was it hard? Speedup?
% \end{itemize}

\newpage
\section{Introduction}

Graphics Processing Units (GPUs) is a technology tailored to perform image processing as fast as possible. They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying speed increases of an order of magnitude compared to similar CPU designs. Initially, GPUs may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another. While this is true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.

Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome. The MVDR technique is known for its notable potential for improving image quality over simple conventional methods, but it is also far more computationally complex. This is because a spatial covariance matrix must be estimated and inverted for each image pixel, the former dominating for systems with less than 20-30 channels, and the latter otherwise.

We propose three different solutions to the complexity issue depending on system size. For smaller systems we use the standard time-domain MVDR method, arithmetically optimize the estimation step, and show that image creation rates of more than 1\,Mpx/s can be achieved when this algorithm is run on a GPU. For medium sized systems we suggest applying the MVDR in the beamspace (frequency domain) instead, which allows for comparable processing rates at a negligible reduction in image quality. Finally, for large systems the estimation and inversion step can be avoided completely by utilising the Low Complexity Adaptive (LCA) beamformer.\todo{disse tre første avsnittene burde sikkert vært kortet ned mye men...}
% 
% 
% The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
% \\
% << MOTIVATION >>
% \\

Related work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\

This article is outlined as follows: First we will present the mathematical formulation and evaluate complexity aspects of the MVDR method operated in time or frequency domain, and on the LCA beamformer. Then we mention some of our experiences with implementing the MVDR on a GPU, before concluding with some results and final thoughts.

\section{Methods}\label{methods}

Suppose that a backscattered wavefield is recorded by an $M$ element uniform linear array, that a suitable set of delays are applied to the channels for the pixel of interest, and let the delayed output from the $m$'th channel be given as $x_m$. By definition a beamformer's output $z[n]$ at time instant $n$ is a weighted sum all the delayed data outputs:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}. However, there exists a class of adaptive beamformers that target this limitation by adapting the weights to the impinging wavefield. The MVDR method is one such technique.

\subsection{MVDR beamforming in the time domain}

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). }\label{mvdr_beamforming}
\end{figure}
The basic idea behind the MVDR method is to compute the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signals coming from the pixel of interest. When the data is delayed this can be formulated as follows:
\begin{gather}
\argmin{\w}\ E\Big\{\big|\w\H[n]\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w\H[n]\1 = 1. \label{mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
This is a convex optimization problem with the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$. In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.

To arrive at a formulation of how $\eR$ can be estimated, we first split the array into subarrays of length $L$ that overlap on all but one sensor. If we let $x_l[n]$ represent the datavector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T.
\end{gather}
then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
\begin{gather}
\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{N_L-1}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays. In the actual design a fraction of the total power of $\eR[n]$ is finally added to its diagonal~\cite{Synnevag2007}, but this step will be ignored here due to its negligible contribution to runtime.

To summarize as in Figure \ref{mvdr_beamforming}: For every pixel in the image we estimate $\eR$ (\ref{spatialR}), solve the linear equation $\eRi\1$ (\ref{weights}), the finally substitute some to get the beamformer output (\ref{mvdr_def}-\ref{weights}).

@@

If (\ref{spatialR}) were to be implemented directly, we note that the estimation step would have a complexity of O($N_K N_L L^2$). By comparison, performing a Gauss Jordan inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L = N_K(M-L+1) \gg L$ the estimation step will dominate. This tends to be true in active systems because we often want to set $L<M/2$ and $N_K\ge>3$ for the MVDR to be robust enough to yield the expected results.

A few properties let us improve these matters notably. First, we may infer from (\ref{spatialR}) that:
\begin{itemize}
\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l\H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one of their triagonals.
\item When summing $R_l[n]$ most of the multiplications need only be computed only once, and most sums can be updated iteratively.
\end{itemize}
Second, each of the diagonals of $R_l[n]\ \forall\ l$ can be computed independently, which is good news for our GPU design.

\subsection{MVDR beamforming in the frequency domain}

@@

The beamspace MVDR formulation very closely resembles that of the time domain MVDR. The main difference is the initial need to Fourier transform the data:
\begin{gather}
\x_{\text{BS},l}[n] = \B\x_l[n], \qquad \B\in \mathbb{C}^{N_b,L}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}.\label{beamspace}
\end{gather}
Here $\B$ is referred to as the Butler matrix, $l$ is the subarray index, and $N_b$ is the number of beams that should be computed. Note has once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{spatialR}), $\w_\text{BS}$ as in (\ref{weights}), and $z$ as in (\ref{z}).

@@

The Low Complexity Adaptive (LCA) beamformer is another take on the optimization problem in (\ref{mvdr_def}). Instead of calculating the weights like in (\ref{weights}), it simply computes the beamformer output for each window in a larger set of windows, and finally applies the one that best fulfils the given optimization criterion. If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as~\cite{Synnevag2008}:
\begin{gather}
\argmin{p}\ E\Big\{\big|\w_p\H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p\H\1 = 1. \label{lca_def}
\end{gather}
Note the similarity between (\ref{mvdr_def}) and (\ref{lca_def}). The LCA and MVDR uses essentially the same optimization criterion, but whereas MVDR searches a continuous solution space for the best set of weights to apply, the LCA beamformer only have $P$ solutions to select from. However, we have found that once the LCA has 10-20 well designed windows available the quality of the images it produces quickly approaches that of the MVDR beamformer.

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
\caption{MVDR implementation.}\label{mvdr_implementation}
\end{figure}

Our GPU implementation consists of three main kernels, one for each of the computing steps illustrated in Figure \ref{mvdr_implementation}. In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$. All pixels are processed by one step before we move on to the next step. Our efforts here were mainly focused on estimating $\eR$, because a highly optimised batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.

>>>>

%&header
% !TeX root = 1_introduction

\input{subheader}
\endofdump

\ifRootBuild\else
  \input{glossary}
  \makeglossaries
\fi

%\beforeepigraphskip

\begin{document}

\newcommand\chapref[2]{\hyperref[#1]{#2.\ref{#1}}}

\chapter{Background}

{

% \newcommand\mat[1]{\boldsymbol{#1}}
\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
\renewcommand*\eR{\mat{\hat R}}
\renewcommand\1{\vec 1}

\section{Introduction}

% Graphics Processing Units (GPUs) is a technology tailored to perform image processing as fast as possible. They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying speed increases of an order of magnitude compared to similar CPU designs. Initially, GPUs may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another. While this is true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.
% 
% Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome. The MVDR technique is known for its notable potential for improving image quality over simple conventional methods, but it is also far more computationally complex. This is because a spatial covariance matrix must be estimated and inverted for each image pixel, the former dominating for systems with less than 20-30 channels, and the latter otherwise. 
% 
% We propose three different solutions to the complexity issue depending on system size. For smaller systems we use the standard time-domain MVDR method, arithmetically optimize the estimation step, and show that image creation rates of more than 1\,Mpx/s can be achieved when this algorithm is run on a GPU. For medium sized systems we suggest applying the MVDR in the beamspace (frequency domain) instead, which allows for comparable processing rates at a negligible reduction in image quality. Finally, for large systems the estimation and inversion step can be avoided completely by utilising the Low Complexity Adaptive (LCA) beamformer.\todo{disse tre første avsnittene burde sikkert vært kortet ned mye men...}
% 
% 
% The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
% \\
% << MOTIVATION >>
% \\

% Related work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\

Waves are made of endless tiny energy transactions, a dance of particles effecting each other in a search for equilibrium. They are caused by disturbances in the forces of nature. The drop of a rock into flat water or rapid displacement of still air cause mechanical waves, while synchronized oscillations of electric and magnetic fields cause electromagnetic waves. Regardless of whether the waves are mechanical or electromagnetic, move perpendicular or parallel to the axis of propagation, what medium they propagate through or the means by which they do so, they are---perhaps surprisingly---all governed by a partial differential equation known as the \emph{wave equation}:
%
\begin{align}
\Delta f = 
\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2} = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
\end{align}
%
where $\Delta$ is the Laplace operator, $f$ is field strength and $c\in\mathbb{R}^+$ is wave propagation speed. If $f$ represents the displacement of a vibrating string, the equation states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side). In acoustics $f$ represents sound pressure, but since the mathematics are all the same we will simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.

Suppose an array of sensors has a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length). A signal source is located at a point $\udot{S}$. We can represent the field observed at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,

An introduction to such notation can be found in \chapref{IV_methods}{IV}.

reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length)

%
\begin{align}
f_\textrm{plane}(\dvec{p}^A,t)  &= A e^{j(\omega t - \dvec{k}^T \dvec{p}^A)} \\
f_\textrm{spheric}(\dvec{p}^A,t) &= \frac{A}{\lVert \dvec{p}^A-\dvec{p}_{AS}^A \rVert} e^{j(\omega t - \dvec{k}^T \lVert \dvec{p}^A-\dvec{p}_{AS}^A \rVert)}
\end{align}
%
% \hyperref[IV_methods]{\chapterref{IV_methods}.\sectionref{IV_methods}}
% \renewcommand{\thesection}{\thechapter.\arabic{section}}

% This article is outlined as follows: First we will present the mathematical formulation and evaluate complexity aspects of the MVDR method operated in time or frequency domain, and on the LCA beamformer. Then we mention some of our experiences with implementing the MVDR on a GPU, before concluding with some results and final thoughts.

\section{Image reconstruction}\label{methods}

Consider a sonar imaging scenario where an encoded signal is transmitted to highlight a surface of interest, and assume that the backscattered wavefield is properly sampled using a uniform linear array of receivers. After stripping the carrier frequency by heterodyning and the signal signature by  matched filtering (Fig. \ref{I_beamforming}), an image can then be formed by coherently combining the receiver outputs to focus at one angle (or point) at the time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

\subsection{Beamforming}
% \ifPhdDoc
% \begin{figure}[htbp]\centering
% \includegraphics[width=\linewidth]{gfx/buske1\figPostfix.eps}
% \else
% \ifPeerReview
% \begin{figure}[htbp]\centering
% \includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
% \else
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/beamforming.eps}
% \fi\fi%
% \caption{Beamforming principle. Signal signature is first removed by matched filtering. Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{I_beamforming}
% \end{figure}

Suppose we record a wavefield with $M$ sensors arranged in a uniformly spaced linear array (ULA) and attach a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length). Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$. We simply notation by making the dependence on $\theta$ implicit from now on.

The beamformer output $z[n]\in\mathbb{C}$ is by definition the weighted sum of all the delayed data samples:
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{2_eq_z}
\end{align}
%
where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}. However, there exists a class of adaptive beamformers that target this limitation by adapting the weights to the impinging wavefield. The MVDR method is one such technique.

\subsection{Time-domain MVDR}

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{2_eq_weights}). }\label{mvdr_beamforming}
\end{figure}
%
The basic idea behind the MVDR method is to compute the set of weights that minimizes noise and signal power accumulated by the array, while ensuring unity gain for signals coming from the pixel of interest. When the data is delayed this can be formulated as follows:
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{1} = 1. \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\dvec{1}}{\dvec{1}^T\Ri[n]\dvec{1}},\label{2_eq_weights}
\end{gather}
%
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x^H[n]\}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$. In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.

To arrive at a formulation of how $\eR$ can be estimated, we first split the array into subarrays of length $L$ that overlap on all but one sensor. If we let $x_l[n]$ represent the datavector from subarray $l$,
%
\begin{gather}
\dvec{x}_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}^T.
\end{gather}
%
then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
%
\begin{gather}
\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{N_L-1}\sumb{n'=n-K}{n+K} \x_l[n]\x_l^H[n] \in\mathbb{C}^{L,L},\label{2_eq_spatialR}
\end{gather}
%
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays. In the actual design a fraction of the total power of $\eR[n]$ is finally added to its diagonal~\cite{Synnevag2007}, but this step will be ignored here due to its negligible contribution to runtime.

To summarize as in Figure \ref{mvdr_beamforming}: For every pixel in the image we estimate $\eR$ (\ref{2_eq_spatialR}), solve the linear equation $\eRi\1$ (\ref{2_eq_weights}), the finally substitute some to get the beamformer output (\ref{2_eq_mvdr_def}-\ref{2_eq_weights}).

@@

If (\ref{2_eq_spatialR}) were to be implemented directly, we note that the estimation step would have a complexity of O($N_K N_L L^2$). By comparison, performing a Gauss Jordan inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L = N_K(M-L+1) \gg L$ the estimation step will dominate. This tends to be true in active systems because we often want to set $L<M/2$ and $N_K\ge>3$ for the MVDR to be robust enough to yield the expected results.

A few properties let us improve these matters notably. First, we may infer from (\ref{2_eq_spatialR}) that:
%
\begin{itemize}
\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l^H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one of their triagonals.
\item When summing $R_l[n]$ most of the multiplications need only be computed only once, and most sums can be updated iteratively.
\end{itemize}
%
Second, each of the diagonals of $R_l[n]\ \forall\ l$ can be computed independently, which is good news for our GPU design.

\subsection{Frequency-domain MVDR}

@@

The beamspace MVDR formulation very closely resembles that of the time domain MVDR. The main difference is the initial need to Fourier transform the data:
%
\begin{gather}
\x_{\text{BS},l}[n] = \mat{B}\x_l[n], \qquad \mat{B}\in \mathbb{C}^{N_b,L}, \qquad \text{where} \qquad [\mat{B}]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}.\label{2_eq_beamspace}
\end{gather}
%
Here $\mat{B}$ is referred to as the Butler matrix, $l$ is the subarray index, and $N_b$ is the number of beams that should be computed. Note has once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{2_eq_spatialR}), $\w_\text{BS}$ as in (\ref{2_eq_weights}), and $z$ as in (\ref{z}).

@@

The Low Complexity Adaptive (LCA) beamformer is another take on the optimization problem in (\ref{2_eq_mvdr_def}). Instead of calculating the weights like in (\ref{2_eq_weights}), it simply computes the beamformer output for each window in a larger set of windows, and finally applies the one that best fulfils the given optimization criterion. If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as~\cite{Synnevag2008}:
\begin{gather}
\argmin{p}\ E\Big\{\big|\w_p^H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p^H\1 = 1. \label{2_eq_lca_def}
\end{gather}
Note the similarity between (\ref{2_eq_mvdr_def}) and (\ref{2_eq_lca_def}). The LCA and MVDR uses essentially the same optimization criterion, but whereas MVDR searches a continuous solution space for the best set of weights to apply, the LCA beamformer only have $P$ solutions to select from. However, we have found that once the LCA has 10-20 well designed windows available the quality of the images it produces quickly approaches that of the MVDR beamformer.

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
\caption{MVDR implementation.}\label{2_eq_mvdr_implementation}
\end{figure}

Our GPU implementation consists of three main kernels, one for each of the computing steps illustrated in Figure \ref{2_eq_mvdr_implementation}. In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$. All pixels are processed by one step before we move on to the next step. Our efforts here were mainly focused on estimating $\eR$, because a highly optimised batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.
